\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements

### 2015-07-28 RITA2015/main.tex
Compressed version -- 29 Pages.

\documentclass{RITA}
\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage[brazilian]{babel}

% Our packages
\usepackage{xspace}
\usepackage{amsmath}
\usepackage[thmmarks,amsmath]{ntheorem}
\usepackage{graphicx}

% Our definitions
\newcommand{\openbox}{\leavevmode
  \hbox to.77778em{%
  \hfil\vrule
  \vbox to.675em{\hrule width.6em\vfil\hrule}%
  \vrule\hfil}}

\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
\newtheorem{algo}{Algorithm}

\theoremstyle{plain}
\theoremheaderfont{\normalfont\bfseries}
\theorembodyfont{\normalfont}
\theoremseparator{}
\theoremindent0cm
\theoremnumbering{arabic}
\theoremsymbol{\ensuremath{\openbox}} 
\newtheorem{example}{Example}

@@

\newcounter{numberInTrivlist}

\newenvironment{numtrivlist}{\begin{list}{\rm \arabic{numberInTrivlist})}
                                         {\usecounter{numberInTrivlist}
                                          \setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

\newenvironment{itemizedTrivlist}{\begin{list}{\rm ~\hspace{2mm} $\bullet$\ }
                                         {\setlength{\leftmargin}{0pt}
                                          \setlength{\rightmargin}{0pt}
                                          \setlength{\itemindent}{12pt}
                                          \setlength{\listparindent}{0pt}}}
                            {\end{list}}

@@

\title{\pisodm: Building Service-Oriented Applications in the Presence of Non-Functional Requirements}
\author{
Khalid Belhajjame
\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
  \texttt{\{kbelhajj@googlemail.com\}}}
\and
Valeria~de~Castro
\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
  \texttt{\{valeria.decastro@urjc.es\}}}
\and
Umberto~Souza~da~Costa
\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
\and
Javier~A.~Espinosa-Oviedo
\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
\and
Martin~A.~Musicante
\footnotemark[3]
\and
Pl\'acido~A.~Souza~Neto
\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
  \texttt{\{placido.neto@ifrn.edu.br\}}}  
\and
Genoveva~Vargas-Solar
\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
  \texttt{\{genoveva.vargas@imag.fr\}}}  
\and  
Jos\'e-Luis~Zechinelli-Martini
\footnotemark[4]
}
\RITAvolume{VIII}
\RITAnumber{1}
\RITAyear{2001}

@@

\begin{abstractinenglish}
Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
The late inclusion of NFRs in the development may compromise the quality of the deployed application.
This paper presents \pisodm, a method and associated tools that
\textit{(i)}  allows the early specification of non-functional requirements in a principled way: users are abstracted away from low level details;
\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
The pro\-po\-sed solution has been utilized in the context of an industrial and real case study.
\end{abstractinenglish}

@@

%Service oriented computing is at the origin of an evolution in the field of software development.
%Service oriented methods advocates for the construction of software systems formed by the composition of heterogeneous, loosely coupled modules.
%These modules (or services) communicate in order to achieve a common purpose.

%(Moreover, IT systems need to evolve according to the business needs.)
%Thus, organizations are seeking for mechanisms to bridge the gap between the actually developed systems and their business needs~\cite{bell}.

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effect produced by the system when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

Functional properties are not the only aspect to be considered in the software development process.
Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed to fit in the application.
%Adding non-functional properties and respecting services constraints while composing services is a complex task that implies implementing protocols for instance authentication protocols to call a service, and atomicity (exception handling and recovery) for ensuring a true synchronization of the results produced by the service methods calls.
%Even if service-oriented computing benefits from reuse, this reuse is usually guided only by functional requirements.
%
%Furthermore,  often they are  not fully considered in the specification process of existing services' oriented development methods. Rather,
%They   are either ensured by the underlying execution platform, or they are programmed through ad-hoc protocols once the application has been programmed.
Ideally, Non-Functional Requirements (NFR) should be considered in every phase of the software development.
Yet, they are partially or rarely methodologically derived from the specification, being usually added once the code has been im\-ple\-men\-ted.
In the context of service oriented computing, the late consideration of non-functional requirements in the development process does not fully preserve the compliance and re\-use expectations promoted by the service oriented paradigm.

%Yet, non-functional properties of  services, often expressed as requirements and constraints in general purpose methodologies, are not usually considered from the early phases of the (service) software process.
%Most methods integrate them only after the application has been implemented.
%This leads to service based applications that are partialy specified and, thereby, partialy compliant with the requirements of the application.

% song title disseminated in the walls of the user's Facebook and Twitter accounts.

%The adoption of non-functional specifications from the early states of development can help the developer to produce applications that  can deal with the application context.

The literature stresses the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches maintain that the convergence of model-driven soft\-ware development, service orientation,   and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review study concludes that current development methods using model-driven methodologies seldom deal with non-functional requirements~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

\bigskip
Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
\pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller} approach.
The goals of  \pisodm are to:
%SOD-M does not provide support for the specification of non-functional requirements, such as
%security, reliability, and efficiency.
\begin{trivlist}
\item \textit{(i)} Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
\item \textit{(ii)} Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
\item \textit{(iii)} Provide a general way of describing non-functional requirements.
\end{trivlist}

@@

\bigskip
This paper is organized as follows.
Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models of the \pisodm and transformation rules.
% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.

\section{Related Work}
\label{sec:relworks}
While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called non-functional properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.

Most research  efforts on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
In the case of service-based applications, non-functional requirements are related to the application itself as well as to its component services.

In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three points of view:
\textit{service level}, \textit{system level} and \textit{business level}.
%In~\cite{Babamir2010} NFRs are denoted as \textit{quality constraints}, which are expressed as logic formulae.
%In~\cite{Yeom2006} authors classify NFRs into \textit{category}, \textit{sub-category} and \textit{property}. Categories include \textit{business}, \textit{service} and \textit{system}.
%Possible \textit{sub-categories} are \textit{security}, \textit{value} or  \textit{interoperability}. The work also defines a \textit{web service quality model}, which considers non-functional properties.
In~\cite{XiaoCZBOLH08} the authors use the terms
\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
The notion of non-functional attribute is used to describe NFRs of the abstract process model.
In the lower level, the composition is annotated with non-functional attributes.

@@

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
A non-func\-tion\-al concern is a general term used to describe non-functional requirements, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

@@

Despite of the different notations found in the literature for classifying NFRs, some non-functional requirements are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
However, distinct hierarchies and models are proposed for NFRs, according to different points of view.
We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to non-functional requirements~\cite{Ameller201542,Ameller2010}. As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with non-functional requirements and gives support to real needs, as presented in a company related to the risk assessment industry.

@@

The proposal in~\cite{ThissenW06} presents steps for  selecting services
by taking QoS information into account. The proposed steps are:
\textit{(i)} identification of relevant QoS information;
\textit{(ii)} identification of basic composition patterns and
QoS aggregation rules for these patterns; and
\textit{(iii)} definition of a selection mechanism of services.
The QoS properties considered are \textit{performance}, \textit{cost}, \textit{reliability} and
\textit{availability}.

Karunamurthy et al.~\cite{Karunamurthy2012787} use the term \textit{non-function parameters} to define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}.
The \textit{Non-Func\-tion\-al Specification Language} (NFSL) is proposed as a domain specific language (DSL) to express \textit{non-function parameters}.

Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe non-functional requirements such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.

Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are intended to be specified as constraints.

%\bigskip
%*****OJO**** QUITAR!! ESTA DISTINCION NO APARECE EN REALIDAD EN EL PAPER NO???, YO QUITARIA SIN MAS ESTE UTLIMO PARRAFO SALVO QUE LE VEAN SENTIDO****
%Most works agree on distinguishing three points of view, namely the point of view of the organization (\textit{Business} view), of the individual service providers (\textit{Service} view) and of the composition designer (\textit{System} view).
%The Business view is concerned with the business logic (\textit{i.e.}, an abstract level of tasks, defined by the guidelines and constraints imposed by the organization).
%Service and System views guides the implementation of the software solution: The Service level is related to the building blocks of the application.
%It may use web services provided by third party sources.
%The System level deals with the coordination of services, to implement the business logic.
%Our proposal, shown in the next section, follows these ideas.

\section{Modeling reliable service compositions with \pisodm}\label{sec:motivation}

In this section we present \pisodm, an MDD-based method for building service compositions with non-functional requirements.
Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.

In the next subsections we present the me\-ta-models from higher to lower levels of abstraction.
The presentation is exemplified by a running example.
The transformations between models are presented in section~\ref{sec:mmrules}.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{./figures/piSOD-M_process.pdf}
\caption{\pisodm Overview.}
\label{fig:piSOD-M}
\end{figure*}

@@

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{E$^3$value meta-model}
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The e$^3$value model of an application identifies the value/information exchange between components of the system.
It is a business model that represents a business case graphically as a set of value exchanges\footnote{See Figure~\ref{fig:CIM:tpme3v}.} ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node (see Figure~\ref{fig:CIM:tpme3v}).
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

\begin{figure*}
\center
\includegraphics[width=0.75\textwidth]{./figures/e3value.pdf}
\hspace*{5cm}\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure*}

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:
\begin{itemize}
\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
\begin{itemize}
%\item{\sf\small get-Status ( usedid ): String ;
\item {\sf\small update-Status ( userid, new-status ): String};
\end{itemize}
\end{itemize}

@@

\subsubsection{BPMN meta-model}

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\end{example}
%
\begin{figure}
\center
\includegraphics[width=0.45\textwidth]{./figures/SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

@@

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-UseCase} meta-model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

>>>>

\documentclass{RITA}
\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage[english]{babel}

% Our packages and definitions
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{paralist} % for compact lists: compactitem, compactenum, compactdesc
\usepackage{lipsum}  % for the example environment
\usepackage{amssymb} % for the \blacksquare symbol
\usepackage{subfigure}

\newcounter{examplecounter}
\newenvironment{example}{%\begin{quote}%
    \refstepcounter{examplecounter}%
  \textbf{Example \arabic{examplecounter}}%
  \quad
}{%
%\end{quote}%
$\blacksquare$ 
}

\let\OLDthebibliography\thebibliography
\renewcommand\thebibliography[1]{
  \OLDthebibliography{#1}
  \setlength{\parskip}{0pt}
  \setlength{\itemsep}{0pt plus 0.7ex}
}

@@

@@

\title{\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements}
\author{
Khalid Belhajjame
\footnote{Universit\'e de Paris - Dauphine -- Paris, France\\
  \texttt{\{kbelhajj@googlemail.com\}}}
\and
Valeria~de~Castro
\footnote{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
  \texttt{\{valeria.decastro@urjc.es\}}}
\and
Umberto~Souza~da~Costa
\footnote{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
  \texttt{\{\{umberto,mam\}@dimap.ufrn.br\}}}
\and
Javier~A.~Espinosa-Oviedo
\footnote{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\
  \texttt{\{javiera.espinosa@gmail.com, joseluis.zechinelli@udlap.mx\}}}  
\and
Martin~A.~Musicante
\footnotemark[3]
\and
Pl\'acido~A.~Souza~Neto
\footnote{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\
  \texttt{\{placido.neto@ifrn.edu.br\}}}  
\and
Genoveva~Vargas-Solar
\footnote{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\
  \texttt{\{genoveva.vargas@imag.fr\}}}  
\and  
Jos\'e-Luis~Zechinelli-Martini
\footnotemark[4]
}
\RITAvolume{22}
\RITAnumber{2}
\RITAyear{2015}

@@

\begin{abstractinenglish}
Specifying non-functional requirements (NFRs) is a complex task, being usually dealt with on the later phases of the software process.
The late inclusion of NFRs in the development may compromise the quality of the deployed application.
This paper presents \pisodm, a method and associated tools that
\textit{(i)}  allows the early specification of NFRs in a principled way: users are abstracted away from low level details;
\textit{(ii)} embraces the MDA philosophy, generating models (code) whenever possible.
Our solution has been used in the context of an industrial and real case study.
\end{abstractinenglish}

@@

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effects it produces when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

Functional properties are not the only aspect to be considered in the software development process.
Non-functional properties, such as data privacy, exception handling, atomicity  and, data persistence, need to be addressed to fit in the application.
Ideally, Non-Functional Requirements (NFR) should be considered in every phase of the software development.
Yet, they are partially or rarely methodologically derived from the specification, being usually added once the code has been im\-ple\-men\-ted.
In the context of service oriented computing, the late consideration of non-functional requirements in the development process does not fully preserve the compliance and re\-use expectations promoted by the service oriented paradigm.

The literature emphasizes the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches assert that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

Our work introduces $\pi$-\textit{Service Oriented Development Method} (\pisodm) to support non-functional aspects of service-oriented applications, taking into account both functional and non-functional requirements, early in the software development process.
Our proposal follows the MDD guidelines and proposes models, practices and techniques for the development of service-based applications.  
\pisodm proposes  the use of \textit{models} to specify a software system at different levels of abstraction.
Models are organized according to the guidelines of the Model Driven Archi\-tec\-ture (MDA)~\cite{miller}.
The goals of  \pisodm are to:

\begin{compactenum}[i]
\item Improve the development process by providing an abstract view of the application and aiding to ensure the conformance to its specification.
\item Reduce the programming effort through the semi-automatic generation of  models for the application, in order to produce concrete implementations from higher-level models.
\item Provide a general way of describing non-functional requirements.
\end{compactenum}

@@

This paper is organized as follows.
Sec\-tion \ref{sec:relworks} summarizes the general principles of existing works for addressing NFP and associating them to service compositions.
Sections~\ref{sec:motivation} and~\ref{sec:mmrules} introduce respectively the meta-models and transformation rules of \pisodm.
% Section~\ref{sec:implementation} describes the \pisodm environment that provides tools for building service based applications according to the \pisodm methodology.
Section~\ref{sec:flyingPig} describes the \FlyingPig case study that we develop for validating our method and discusses lessons learned.
Finally, Section~\ref{sec:conclusions} concludes the paper and gives research perspectives.

\section{Related Work}
\label{sec:relworks}
While Functional Requirements establish \textit{what} is computed by an application, Non-Functional Requirements (NFRs) are concerned with \textit{how} the task is performed.
NFRs include aspects such as performance, authentication and quality constraints.
These requirements are usually specified by conditions, called non-functional properties.
Non-functional properties are also referred to as constraints, quality attributes, quality goals, quality of service requirements and non-behavioral requirements~\cite{Chung91,MylopoulosBook99,Chung2009}.
%
Most research  on NFRs focus on the evaluation of compliance by the software system as a whole. %~\cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.
In service-based applications, NFRs are related to the application itself as well as to  its component services.

In~\cite{Babamir2010,Yeom2006} non-functional properties of web services are classified according to three perspectives:
\textit{service}, \textit{system} and \textit{business levels}.
In~\cite{XiaoCZBOLH08}  authors use the terms
\textit{non-functional attributes}, \textit{composition mo\-del}  \textit{entity} and \textit{mo\-del entity}  to classify different concepts related to NFRs.
The notion of non-functional attribute is used to describe NFRs of the abstract process model.
In the lower level, the composition is annotated with non-functional attributes.

@@

Schmeling et al.~\cite{SchmelingCM11} present an approach and a toolkit for specifying and implementing web service compositions with support to several NFRs.
Their approach defines abstraction levels, where the terms \textit{Non-Functional Concerns}, \textit{Non-Functional Actions} and \textit{Non-Functional Attributes} are used at each level (in decreasing order of abstraction).
A non-func\-tion\-al concern is a general term used to describe NFRs, such as  \textit{security}, \textit{reliability} or \textit{transactional behavior}.
Each  concern is refined into a set of non-func\-tion\-al  at\-tri\-butes, where
a non-func\-tion\-al at\-tri\-bute represents some behavior, to be refined into non-functional actions.
For instance, \textit{encryption} is a non-func\-tion\-al  action  which provides the implementation of the non-functional attribute \textit{confidentiality}, which is part of the \textit{security} non-functional concern.
%Non-functional actions related to a common concern are grouped into \textit{non-functional activities}.

@@

Despite of the different notations found in the literature for classifying NFRs, some NFRs are frequently considered, such as \textit{security}, \textit{performance}, \textit{reliability}, \textit{usability}, and \textit{availability}.
However, distinct hierarchies and models are proposed for NFRs, according to different perspectives.
We have identified a number of approaches~\cite{DAmbrogio06,CholletL09,SchmelingCM11,BasinDL06,Fabra2011,OvaskaEHPA10} that use MDD (Model Driven Development) for designing and developing applications.

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}. As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.

@@

The proposal in~\cite{ThissenW06} presents steps to  select services
by taking QoS information into account:
\textit{(i)} identification of relevant QoS information;
\textit{(ii)} identification of basic composition patterns and
QoS aggregation rules for these patterns; and
\textit{(iii)} definition of a selection mechanism of services.
The authors consider  \textit{performance}, \textit{cost}, \textit{reliability} and
\textit{availability}.

Karunamurthy et al.~\cite{Karunamurthy2012787} define NFRs, such as \textit{cost}, \textit{response time}, \textit{availability}, \textit{security}, \textit{reliability} and \textit{reputation}, as  \textit{non-function parameters}.
The \textit{Non-Func\-tion\-al Specification Language} 
%(NFSL)
 is proposed as a domain specific language 
 %(DSL) 
 to express \textit{non-function parameters}.

Liu et al.~\cite{Liu20121080} use the term \textit{QoS parameter} to describe NFRs such as \textit{cost}, \textit{execution duration}, \textit{accuracy}, \textit{security}, \textit{integrity}, \textit{availability} and \textit{reliability}.
In the same way, Tran et al.~\cite{Tran2012531} use the term \textit{QoS policies} to classify similar non-functional requirements.

Li et al.~\cite{Li2013} associate \textit{dimensions} to  \textit{QoS parameters} to classify NFRs.
For instance, the \textit{time} dimension is associated to the \textit{execution time} and \textit{communication time} parameters; the \textit{spatial} dimension is associated to the \textit{storage capacity} and \textit{message length} parameters; the \textit{reliability} dimension is associated to the \textit{availability} and \textit{reliability} parameters and the \textit{cost} dimension is associated to the \textit{service cost} parameter.
Rumpel et al.~\cite{Rumpel2012}  associate \textit{quality requirements} to  \textit{quality properties}. Quality requirements are to be specified as constraints.

\section{Modeling Reliable Service Compositions with \pisodm}\label{sec:motivation}

In this section we present \pisodm, an MDD-based method for building service compositions with NFRs.
Our work provides meta-models for modeling functional and non-functional requirements organized in three levels (Figure~\ref{fig:piSOD-M}): CIM (\textit{Computational Independent Models}), PIM (\textit{Plat\-form Independent Models}) and PSM (\textit{Platform Specific Models}).
Given  high-level models specified at the CIM level, \pisodm proposes the semi-automatic refinement of these models, for the generation of a set of models at the other levels of abstraction.
The refinement process is driven by transformation rules specified between the meta-models.
%
The me\-ta-models are presented from higher to lower levels of abstraction along with a running example.
The transformations between models are presented in section~\ref{sec:mmrules}.

\begin{figure*}[h]
\centering
\includegraphics[width=0.75\textwidth]{./figures/piSOD-M_process.pdf}
\caption{\pisodm Overview.}
\label{fig:piSOD-M}
\end{figure*}

@@

% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{E$^3$value Meta-Model}
% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .

The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node.
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

\begin{figure*}[h]
\center
\subfigure{\includegraphics[width=0.55\textwidth]{./figures/e3value.pdf}}
~ 
\subfigure{\includegraphics[width=0.4\textwidth]{./figures/3ValueKey.pdf}}
\caption{\label{fig:CIM:tpme3v} e3value model for ``To Publish Music''.}
\end{figure*}

\begin{example}[To Publish Music]\label{ex:toPublicMusic}
Let us consider the scenario ``To Publish Music'', used as a running example in this section:
An organization wants to provide the service based application ``To Publish Music'' that monitors the music listened by a user during some periods of time and sends the song title  to this person's Twitter and Facebook accounts.
In this way, the user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
The application is based on three external actors ({\em Spotify, Twitter} and {\em Facebook}).
The following (external) services will be used by the application:

\begin{compactitem}
\item The  service   Spotify exports a meth\-od for obtaining information  about the music a given user is listening:
 {\sf\small get-Last-Song ( userid ): String}; 
\item The services Facebook and Twitter export meth\-ods for  updating the status of a given user:
 {\sf\small update-Status ( userid, new-status ): String};
\end{compactitem}

@@

\subsubsection{BPMN Meta-Model}

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusicBPMN}
Figure \ref{fig:CIM:tpmbpmn} shows the BPMN model\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario.
It starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}).
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\end{example}
%
\begin{figure}[h]
\center
\includegraphics[width=0.4\textwidth]{./figures/SC.pdf}
\caption{\label{fig:CIM:tpmbpmn} BPMN model for ``To Publish Music''.}
\end{figure}

@@

 % .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
\subsubsection{\textit{$\pi$-UseCase} Meta-Model}% .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .  .   .
